---
title: "Formulas"
output: html_notebook
---
Often, the data we get does not tell the story. Often, there are other issues we must deal with. For instance: how has the data changed over time? How do we level the playing field between different populations? 

One of the most common data analysis techniques is to look at change over time. The most common way of comparing change over time is through percent change. The math behind calculating percent change is very simple, and you should know it off the top of your head. The easy way to remember it is:

`(new - old) / old` 

Or new minus old divided by old. Your new number minus the old number, the result of which is divided by the old number. To do that in R, we can use `dplyr` and `mutate` to calculate new metrics in a new field using existing fields of data. 

So first we'll import dyplr.
```{r}
library(dplyr)
```
Now we'll import a common and simple dataset of population estimates for every county in the US. The estimates data has data from 2010 to 2016. 
```{r}
population <- read.csv("../Data/population.csv")
head(population)
```
The code to calculate percent change is pretty simple. Remember, with `summarize`, we used `n()` to count things. With `mutate`, we use very similar syntax to calculate a new value using other values in our dataset. So in this case, we're trying to do (new-old)/old, but we're doing it with fields. If we look at what we got when we did `head`, you'll see there's POPESTIMATE16 as the new data, and we'll use POPESTIMATE2015 as the old data. So we're looking at 1 year. Then, to help us, we'll use arrange again to sort it, so we get the county with the fastest growing population over one year. 
```{r}
population %>% mutate(
  change = (POPESTIMATE2016 - POPESTIMATE2015)/POPESTIMATE2015
) %>% arrange(desc(change))
```
But if we look at change, we'll see that it's a decimal point. That's because for percent change to be a percent, you must multiply it by 100. You do that this way:
```{r}
population %>% mutate(
  change = ((POPESTIMATE2016 - POPESTIMATE2015)/POPESTIMATE2015)*100,
  longchange = ((POPESTIMATE2016 - POPESTIMATE2010)/POPESTIMATE2010)*100
) %>% arrange(longchange)
```
### Assignment

How has Nebraska's electorate changed from 2010 to the election of Donald Trump in 2016? Specifically, how has the total number of voters, Republicans, Democrats and Non-Partisans changed in that time **in each county** in Nebraska? Which counties have changed the most in terms of the number of registered voters? [Here is your dataset](https://www.dropbox.com/s/7epokhv6ruujfgf/registeredvoters.csv?dl=0).

#### Rubric

1. Did you import the data correctly?
2. Did you mutate the data correctly? Did you do it in one step?
3. Did you sort the data correctly?
4. Did you explain each step using Markdown?

### Adjusting for population

Another common formula used in data journalism is adjusting for population. Often we do that by calculating a rate -- X per Y population. A common one is per capita -- per 100,000. 

The most common place you see this is in crime rates. How do you compare the crime in New York City to York, Neb.? On a numeric basis, you can't. There's more people in several blocks of New York than the whole town of York. But you can on a rate basis. 

The formula for a per capita figure is `(THING / POPULATION) * 100000`. So if we're looking at murder or left handed people or violin players -- you divide that by the population of the place you are looking at and multiply it by 100,000.

Without looking, what town has the highest violent crime rate in Nebraska? If you've been here, you'd think Omaha, the state's largest city.

You would be wrong. At least in 2016, the last time the FBI updated their nationwide statistics. So let's take a look. 

```{r}
crime <- read.csv("../Data/NebraskaViolentCrime.csv")
head(crime)
```
Note there are cities with data missing. So we have to use our filtering skills to get rid of those -- we can't calculate a rate if we don't know the population. 

```{r}
crime %>% filter(Population != "")
```
Now we need to apply our formula. In this case, we have a field called `Violent.crime.total` and a field called `Population`. The rest is simple. We'll filter out the bad data, then use `mutate` to create a field called Violent.crime.rate and then, to answer our question, we'll sort it. 

```{r}
crime %>% filter(Population != "") %>%
  mutate(
    Violent.crime.rate = (Violent.crime.total/Population)*100000
  ) %>% arrange(desc(Violent.crime.rate))
```
Beatrice, on a per capita basis, has more violent crime than Omaha. 

### Assignment

Using the same data, which Nebraska city has the highest murder rate per capita?  [Here is your dataset](https://www.dropbox.com/s/l5u7qvvkjczw1w7/NebraskaViolentCrime.csv?dl=0).

#### Rubric

1. Did you import the data correctly?
2. Did you mutate the data correctly? Did you do it in one step?
3. Did you sort the data correctly?
4. Did you explain each step using Markdown?
