---
title: "Joins"
output: html_notebook
---
In databases, when we have two tables that are related to each other by a common element, then our database is called a relational database. When all of our data is in a single file, that's called a flat file. 

Often, in data, we have one set of information stored in a table over here, and another set of information stored in a table over here. At the university, your student records are scattered in tables all over. Somewhere, there is a master student record, that has your name, birthdate, ID number, home address and other basic info. Then, over in the registrars office, we have the classes you took and the grades you received. Over here, we have the bursars office, which shows how much you owe in tuition and how much you've paid. If we wanted to get a single table together that showed how much you paid for each grade you got, we'd have to JOIN them together somehow. 

In dplyr, like SQL, it's called a join. [The library's documentation on this is quite good](https://dplyr.tidyverse.org/reference/join.html), but let's walk through it. I've got two datasets that I want to join together and calculate a new number from something in both. 

Here is the question we're trying to answer: What Nebraska city has seen the largest growth in taxable sales between two years?

First, let's import our library:
```{r}
library(dplyr)
```
Now we'll fetch some data.
```{r}
taxes15 <- read.csv("../Data/taxes2015.csv")
taxes16 <- read.csv("../Data/taxes2016.csv")
head(taxes15)
head(taxes16)
```
Note right away: What format are the Net.Taxable.Sales and Nebraska.Sales.Tax fields? What can we do about that? (Hint: Lesson 2.)
```{r}
clean2015 <- taxes15 %>% mutate(
  NetTaxableSalesNum = as.numeric(gsub(",", "", Net.Taxable.Sales)),
  NebraskaSalesTaxNum = as.numeric(gsub(",", "", Nebraska.Sales.Tax))
)
head(clean2015)

clean2016 <- taxes16 %>% mutate(
  NetTaxableSalesNum = as.numeric(gsub(",", "", Net.Taxable.Sales)),
  NebraskaSalesTaxNum = as.numeric(gsub(",", "", Nebraska.Sales.Tax))
)
head(clean2016)
```
Now with clean numeric data, we're ready to join them. But first we need to understand what we're talking about. 

When you bring two different datasets together, it's rare that you have exactly the same number of records in one as you do in another. Usually when that happens, you have a known number of entities -- 50 states, for example. Or a set cohort of people who all have unique identifiers. More often than not, in my experience, you have a differing number of entities in the two datasets. Entities will drop out -- a state won't report data one year -- or your matching criteria aren't as clean as a unique identifier. 

So a question we have with joins is often: What do we do with the mismatches? Which table is more important? Do we only want matches? Or do we want all of one table and just join the second table where they match? Or the other way around? Or everything regardless of matches? 

The answers determine what join you use. 

If you only want matches, you use what's called an inner join, or sometimes called an equivalency join. The result you get is all the records that match between table 1 and table 2. Mismatches are discarded.

If you want all of the records from table 1 and only records that match from table two, you use what's called a left join, meaning the important table is on the left in the reading left to right style of westerners. 

If you want all the records from table 2 and only records that match from table 1, it's a right join. 

If you want everything from both, matching or not, with blank data in the places where there isn't a match, you want a full join. 

We're going to do an inner join. To calculate which city had the largest growth in taxable sales, we need data from both years. Missing data doesn't help us at all. 

The first step to doing a join is knowing how many rows you are starting with, so you can see what you end up with. If you have data on 50 states and you expect to have 50 states when you are done, you need to check if that's so. If you have 49, you have a problem. 

In R, we use `nrow` to determine the *n*umber of *row*s. 
```{r}
nrow(clean2015)
```
So in 2015, we have data on 458 towns. 
```{r}
nrow(clean2016)
```
And we have 456 towns in 2016. So right away, we know we're going to have two fewer records when we merge them together. But are the same 456 towns in 2016 the 2015 data? Let's find out.

The way joins in dplyr work is that the left table is the first table referenced, and the type of join is what comes next. In it, you must specify which table you are joining, and then you must tell dplyr what fields to use in the join, otherwise it will try all of them and you'll get nothing. So for our purposes, we need to join by County and City. Why not just city? What would happen if we have two towns with the same name in two different places? It's not common, but it does happen. 
```{r}
clean2015 %>% inner_join(clean2016, by=c("County", "City"))
```
And just like that, we now have a table with tax data for 452 towns. So the answer to our question of how many towns in the 2015 data are in the 2016 data and vice versa, it's 452 -- six fewer than 2015. 

Now, with that syntax, we can chain together other things we have learned -- percent change calculations from the last lesson maybe? 
```{r}
clean2015 %>% 
  inner_join(clean2016, by=c("County", "City")) %>%
  mutate(
    Change = ((NetTaxableSalesNum.y-NetTaxableSalesNum.x)/NetTaxableSalesNum.x)*100
  ) %>%
  arrange(desc(Change)) %>%
  select(County, City, Change)
```
Whoa, what happened in Harrisburg in Banner County? As an old editor of mine used to say: I don't know, go do some of that reporting shit you do. 

### Assignment

There are two data files: [frl15](https://www.dropbox.com/s/5h2yauewwfuva73/frl15.csv?dl=0) and [frl16](https://www.dropbox.com/s/hu3035y0ft1q403/frl16.csv?dl=0). They are the Free and Reduced Lunch participation totals for every school in Nebraska. I want you to join them together into a single table and calculate the percent change from 2015 to 2016 and sort them by the largest change. Which school in Nebraska saw the largest increase in participation in free and reduced school lunches, which is a proxy for poverty? 

To complete this assignment, you are going to have to join the dataset together. There is a field called `codistsch` that is a unique identifier for the school. The problem? R is going to interpret it as a number. It is not. So you're going to have to implement [overriding column types](http://r4ds.had.co.nz/import.html), which is not hard.

### Rubric

1. Did you import the data correctly?
2. Did you compare the number of records in the two datasets correctly?
3. Did you use the join syntax correctly?
4. Did you calculate the percent change correctly?
5. Did you document your steps in Markdown?